---
title: "Mapping Oʻahu Extents and Conditions"
author: "Alemarie Ceria"
date: 03/20/2024
format: 
  html: 
    code-overflow: wrap
---

## Desired Output(s)

Quarto website or Shiny Dashboard of extents and conditions of each island by moku

*Current Focus*: Oʻahu Extents and Conditions

## Helper Functions

```{r}
#| label: import-data-from-google-sheets-function
#| code-fold: true
#| code-summary: "Function to Import Extents and Conditions Data from Google Sheets"
import_mhi_extents_and_conditions_gs <- function(url_str) {
  imported_sheets <- gs4_get(url_str) %>%
    pluck("sheets")

  sheet_names_lower <- tolower(imported_sheets$name)

  sheets_data <- purrr::map(imported_sheets$name, ~ {
    Sys.sleep(2) 
    read_sheet(url_str, sheet = .x)
  })

  names(sheets_data) <- sheet_names_lower

  return(sheets_data)
}
```

```{r}
#| label: subset-mhi-dfs-function
#| code-fold: true
#| code-summary: "Function to Subset Main Hawaiian Islands (MHI) Dataframes"
subset_mhi_dfs <- function(mhi_dfs, subset, island_str = NULL, moku_str = NULL, extent_str = NULL) {
  if (!subset %in% c("conditions", "extents")) {
    stop("Subset must be either 'conditions' or 'extents'")
  }
  
  dfs <- mhi_dfs[[subset]]
  
  filtered_dfs <- list()
  
  for (name in names(dfs)) {
    df <- dfs[[name]]
    
    conditions_met <- c(
      is.null(island_str) || grepl(island_str, name, ignore.case = TRUE),
      is.null(moku_str) || grepl(moku_str, name, ignore.case = TRUE),
      is.null(extent_str) || grepl(extent_str, name, ignore.case = TRUE)
    )
    
    if (all(conditions_met)) {
      filtered_dfs[[name]] <- df
    }
  }
  
  return(filtered_dfs)
}
```

```{r}
#| label: export-tidied-dfs-to-csv-function
#| code-fold: true
#| code-summary: "Function to Export Tidied Dataframes to CSV"
export_tidied_dfs_to_csv <- function(df, desc_name, dir_path = "data/tidied/") {
  # Ensure the directory exists; if not, create it
  if (!dir.exists(dir_path)) {
    dir.create(dir_path, recursive = TRUE)
  }
  
  # Construct the file path with the current date and descriptive name
  file_path <- paste0(dir_path,
                      format(Sys.Date(), "%Y-%m-%d"),
                      "_tidied-", desc_name, ".csv")
  
  # Write the dataframe to a CSV file
  write.csv(df, file_path, row.names = FALSE)
  
  # Optionally, print a message indicating where the file was saved
  message("File saved to: ", file_path)
}
```

```{r}
#| label: tidy-oahu-conditions-data-function
#| code-fold: true
#| code-summary: "Function to Tidy Oʻahu Conditions Data"
tidy_conditions_data <- function(df, remove_cols_vect = NULL, remove_rows_vect) {
  df_tidied <- df %>%
  mutate(island = "Oʻahu",
         moku = if_else(ISLAND == "MOKU", Oahu, NA_character_),
         ecosystem_type = if_else(...3 == "Ecosystem Type", Oahu, NA_character_)) %>%
  relocate(island, moku, ecosystem_type, .before = "...1") %>%
  fill(c(moku, ecosystem_type), .direction = "down") %>%
  rename("typology_class" = ...1,
         "descriptor" = ...2,
         "unit" = ...3,
         "Variable Values: Opening Value" = ISLAND,
         "Variable Values: Closing Value" = Oahu,
         "Actively Managed or Conserved" = ...6,
         "Reference Level Values: Upper Level" = ...7,
         "Reference Level Values: Lower Level" = ...8,
         "Indicator Values (Rescaled): Opening Value" = ...9,
         "Indicator Values (Rescaled): Closing Value" = ...10,
         "Indicator Values (Rescaled): Change in Indicator" = ...11) %>% 
    {if (!is.null(remove_cols_vect)) select(., -all_of(remove_cols_vect)) else .} %>% 
  slice(-remove_rows_vect) %>%
  fill(c(typology_class, descriptor, unit), .direction = "down") %>%
  mutate(typology_class = str_replace_all(typology_class, "\\s+|\\n", " ")) %>%
  separate(typology_class, into = c("typology_class_code", "typology"),
           sep = "\\) ",
           extra = "merge",
           fill = "right") %>%
  mutate(
    typology_class_code = str_trim(str_replace(typology_class_code, "\\(", "")),
    typology = str_trim(str_replace(typology, "^\\)", ""))
    ) %>%
  mutate(
    across(where(is.character), ~gsub("([A-Za-z]+)/ ([A-Za-z]+)", "\\1/\\2", .))) %>%
  mutate(across(.cols = 8:15, .fns = as.character)) %>%
  pivot_longer(
    cols = c(8:15),
    names_to = "measurement",
    values_to = "value"
  ) %>%
  separate(measurement, into = c("category", "measurement"), sep = ": ", convert = TRUE) %>%
  mutate(
    category = str_trim(category),
    measurement = str_trim(measurement)
  ) %>%
  mutate(
    category = as.factor(category),
    measurement = as.factor(measurement),
    value = as.numeric(value)
  ) %>%
    mutate(measurement = coalesce(measurement, "Actively Managed or Conserved")) %>% 
    filter(category != "Indicator Values (Rescaled)")
  
    return(df_tidied)
}
```

## Data Importation

### Data Sources

### Procedure

```{r}
#| label: install-packages-and-load-libraries
#| code-fold: true
#| code-summary: "Install Packages and Load Libraries"
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(renv, yaml, rmarkdown, sf, leaflet, mapview, htmlwidgets,
               janitor, tidyverse, googlesheets4, terra, lwgeom,
               viridis, stars)
```

```{r}
#| label: configs
#| code-fold: true
#| code-summary: "Configurations"
gs_urls <- yaml::read_yaml("code/gs_urls.yml")
```

```{r}
#| label: authenticate-google-account
#| code-fold: true
#| code-summary: "Authenticate Google account"
#| eval: false
gs4_auth()
```

::: {.panel-tabset}

#### General

Import extents and conditions data from Google Sheets for MHI:

```{r}
#| label: import-data-from-google-sheets
#| eval: false
mhi_extents_dfs <- import_mhi_extents_and_conditions_gs(gs_urls$extents)

mhi_conditions_dfs <- import_mhi_extents_and_conditions_gs(gs_urls$conditions)
```

Efficiently load extents and conditions data for MHI:

```{r}
#| label: efficiently-load-mhi-extents-and-conditions-data
# Export once or when updated
# mhi_dfs <- mhi_dfs <- list(
#   conditions = mhi_conditions_dfs,
#   extents = mhi_extents_dfs
#   )
# save(mhi_dfs, file = "data/raw/mhi_dfs.RData")

base::load("data/raw/mhi_dfs.RData")
```

List of MHI dataframes:

```{r}
names(mhi_dfs$extents)
names(mhi_dfs$conditions)
```

#### Specific

Subset MHI dataframes:

```{r}
#| label: import-oahu-extents-and-conditions-dat
oahu_extents_dfs <- subset_mhi_dfs(
  mhi_dfs, subset = "extents", island_str = "oahu"
  )

oahu_conditions_dfs <- subset_mhi_dfs(
  mhi_dfs, subset = "conditions", island_str = "oahu"
  )
```

List of Oʻahu dataframes:

```{r}
names(oahu_extents_dfs)
names(oahu_conditions_dfs)
```

:::

## Data Cleaning and Preparation

### Tidying and Merging Raw Datasets

::: {.panel-tabset}

#### Extents

##### Marine

```{r}
#| label: tidy-oahu-extents-data
#| code-fold: true
#| code-summary: "Tidy Oʻahu Marine Extents Data"
new_marine_extent_names <- c("island", "moku", "temp1", "temp2", "temp3", "temp4", "temp5", "temp6")

tidied_oahu_marine_extents_df <- oahu_extents_dfs$marine_oahu %>% 
  select(1, 2, 8:13) %>% 
  set_names(new_marine_extent_names) %>%
  mutate(island = "Oʻahu") %>%
  fill(moku, .direction = "down") %>%
  mutate(
    moku = case_when(
      lag(island) == "MOKU" ~ moku,
      TRUE ~ moku
    )
  ) %>% 
  filter(temp1 %in% c("Area (km2)", "Opening Stock (2004-2014)")) %>%
  pivot_longer(
    cols = c(4:8),
    names_to = "temp",
    values_to = "habitat"
  ) %>% 
  select(-c(3, 4)) %>% 
  mutate(
    opening_stock = if_else(grepl("^\\d+(\\.\\d+)?$|^-$", habitat), as.character(habitat), NA_character_),
    opening_stock = if_else(opening_stock == "-", "0", opening_stock),
    opening_stock = as.numeric(opening_stock)
  ) %>% 
  mutate(row = row_number()) %>%
  # Create a grouping variable for habitats and their corresponding stock values
  group_by(island, moku) %>%
  mutate(group = cumsum(habitat %in% c("Soft Bottom", "Other Hard Bottom", "Rock/ Boulder", "Pavement", "Coral Dominated Hard Bottom") | row == 1)) %>%
  # Shift opening_stock values up within each group
  mutate(opening_stock = if_else(group > 1 & !is.na(opening_stock), opening_stock, lead(opening_stock, n = 5))) %>%
  ungroup() %>%
  # Filter out the stock value rows, leaving only habitat rows
  filter(habitat %in% c("Soft Bottom", "Other Hard Bottom", "Rock/ Boulder", "Pavement", "Coral Dominated Hard Bottom")) %>%
  select(-row, -group) %>% 
  mutate(
    across(c(island, moku), as.factor),
    habitat = as.factor(map_chr(habitat, ~ .x[1]))
    ) %>% 
  # filter(habitat != "Other Hard Bottom") %>% 
  mutate(habitat = if_else(habitat == "Rock/ Boulder", "Rock/Boulder", habitat))

paged_table(tidied_oahu_marine_extents_df)
```
<br>

##### Terrestrial

```{r}
#| eval: false
new_terrestrial_extent_names <- c("island", "moku", "habitat")
tidied_oahu_terrestrial_extent_df <- oahu_extents_dfs$terrestrial_oahu %>% 
  mutate(island = "Oʻahu",
         moku = if_else(TERRESTRIAL == "MOKU", ...2, NA_character_)) %>% 
  fill(c(island, moku), .direction = "down") %>% 
  slice(-c(1:30)) %>% 
  select(-c(1, 12:15))
```

#### Conditions

```{r}
#| label: tidy-oahu-conditions-data
#| code-fold: true
#| code-summary: "Tidy Oʻahu Conditions Data"
tidied_oahu_conditions_dfs_list <- list(
  list(df = oahu_conditions_dfs$waianae_oahu, 
       remove_cols_vect = c(15:20), 
       remove_rows_vect = c(1:6, 18:27, 39:45, 55:59, 62:67, 83:88)),
  list(df = oahu_conditions_dfs$koolaupoko_oahu, 
       remove_rows_vect = c(1:6, 18:24, 36:42, 55:64, 67:74, 90:96, 98:106)),
  list(df = oahu_conditions_dfs$ewa_oahu, 
       remove_rows_vect = c(1:6, 18:23, 35:40, 53:60, 63:71)),
  list(df = oahu_conditions_dfs$kona_oahu, 
       remove_rows_vect = c(1:5, 17:22, 34:39, 52:60, 63:71, 87:92, 94:97)),
  list(df = oahu_conditions_dfs$koolauloa_oahu, 
       remove_rows_vect = c(1:5, 17:22, 34:39, 52:56, 59:64, 80:84, 86:89, 95:103)),
  list(df = oahu_conditions_dfs$waialua_oahu, 
       remove_rows_vect = c(1:5, 17:22, 34:39, 52:56, 59:64, 80:85))
)

tidied_oahu_conditions_df <- map_dfr(tidied_oahu_conditions_dfs_list, ~do.call(tidy_conditions_data, .x)) %>% 
  mutate(moku = gsub("Ewa", "ʻEwa", moku))

paged_table(tidied_oahu_conditions_df)
```

:::

## Data Exportation

### Exporting Processed Data

```{r}
#| label: export-tidied-dfs-to-csv
#| eval: false
tidied_dfs_list <- list(
  list(df = tidied_oahu_conditions_df, desc_name = "oahu-conditions")
#   list(df = tidied_oahu_marine_extents_df, desc_name = "oahu-marine-extents")
)

map(tidied_dfs_list, ~export_tidied_dfs_to_csv(.x$df, .x$desc_name))
```

### Generating Data Dictionaries

## Visualizations

```{r}
#| eval: false
#| echo: false
crs <- "EPSG:4326"
```

```{r}
#| eval: false
#| echo: false
moku <- st_read("data/spatial/moku/Moku_NEW.shp", quiet = TRUE)

# Assuming 'moku' is your spatial vector data and needs to be transformed as well
moku_transformed <- moku %>%
  clean_names() %>%
  mutate(name2 = str_to_title(name2)) %>%
  st_transform(crs = crs)

waianae_moku <- moku_transformed %>% 
  filter(name2 == "Waianae")
```

```{r}
#| eval: false
#| echo: false
mhi_marine_extent <- terra::rast("data/spatial/extent-layers/marine/Benthic_Habitat_5Classes_2022_10.tif") %>% 
  terra::project(crs)

waianae_moku_marine_extent <- mhi_marine_extent %>% 
  crop(waianae_moku) %>% 
  as.factor()

# waianae_moku_marine_extent@ccp$getCategories()
# 
# getCategories()
# activeCat(waianae_moku_marine_extent)

# waianae_moku_marine_extent@cpp$activeCat() <- 1
```

```{r}
#| eval: false
#| echo: false
# Define the colors for the habitat categories that are actually present in the raster
habitat_colors <- c("white", "#F4E8C1", "#808080", "green", "#FF4500")  # Excludes "Other Hard Bottom"

# Update the RAT with the correct values and colors, excluding "Other Hard Bottom"
rat <- data.frame(
  value = c(0, 1, 3, 4, 5),  # Updated to reflect the values that are present
  habitat = c("NA", "Soft Bottom", "Rock/Boulder", "Pavement", "Coral Dominated Hard Bottom"),
  color = habitat_colors
)

# Assign the updated RAT to the raster
levels(waianae_moku_marine_extent) <- rat

na_color <- "black"  
rat$color[is.na(rat$color)] <- na_color
```

```{r}
#| eval: false
#| echo: false
colored_waianae_moku_extents <- rat %>% 
  left_join(tidied_waianae_oahu_extents, ., by = "habitat")
```


```{r}
#| warning: false
#| eval: false
#| echo: false
terra::plot(waianae_moku_marine_extent, col = colored_waianae_moku_extents$color)
lines(waianae_moku, col="forestgreen", lwd=5) 
```

```{r}
#| eval: false
#| echo: false
leaflet() %>%
  setView(lng = -157.99827222584105, lat = 21.47, zoom = 10) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addRasterImage(waianae_moku_marine_extent, opacity = 0.8, colors = rat$color) %>%
  # addLegend(pal = rat$Color, values = rat$Value, opacity = 1, title = "Habitat") %>% 
  addPolygons(data = waianae_moku, color = "#F4C2A5", opacity = 0.5,
              label = ~as.character(name2),
              labelOptions = labelOptions(
                direction = 'auto', 
                style = list('font-size' = '15pt',
                             'font-weight' = 'bold'), 
                noHide = TRUE)
              ) 
```

```{r}
#| eval: false
#| echo: false
# Generate a color palette
palette <- colorFactor(palette = "viridis", domain = waianae_moku_marine_extent_df$Habitat)

# Create the Leaflet map
leaflet(waianae_moku_marine_extent_sf) %>%
  setView(lng = -158.18700955981603, lat = 21.44041625309884, zoom = 9) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addPolygons(fillColor = ~colors(layer), color = ~colors(layer), fillOpacity = 0.7, weight = 0) %>%
  addLegend(pal = colors, values = ~layer, opacity = 1, title = "Habitat") %>% 
  addPolygons(data = waianae_moku, color = "#F4C2A5", opacity = 0.5,
              label = ~as.character(name2),
              labelOptions = labelOptions(
                direction = 'auto', 
                style = list('font-size' = '15pt',
                             'font-weight' = 'bold'), 
                noHide = TRUE)
              ) 
```

```{r}
#| eval: false
#| echo: false
mapview(waianae_moku_marine_extent, legend = TRUE, trim = TRUE, 
        use.layer.names = TRUE) + 
  mapview(waianae_moku, label = waianae_moku$name2)
```

```{r}
#| code-fold: true
#| code-summary: "Create Interactive Map"
#| fig-height: 10
#| fig-width: 15
#| eval: false
#| echo: false

leaflet() %>%
  setView(lng = -158.18700955981603, lat = 21.44041625309884, zoom = 12) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addTiles(urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/World_Imagery/MapServer/tile/{z}/{y}/{x}",
           options = tileOptions(opacity = 0.9)) %>%
  addTiles(urlTemplate = "https://server.arcgisonline.com/ArcGIS/rest/services/Reference/World_Boundaries_and_Places/MapServer/tile/{z}/{y}/{x}",
           options = tileOptions(opacity = 0.9)) %>%
    addRasterImage(marine_extent, opacity = 0.8, colors = viridis::viridis(100)) %>%
  addPolygons(data = moku_transformed, 
              color = "#F4C2A5",
              opacity = 0.5,
              label = ~as.character(name2), # Add hover labels
              labelOptions = labelOptions(
                direction = 'top', 
                style = list('font-size' = '20pt'),
                noHide = TRUE
                )
              )
```

## Resources

- [Esri | Sentinel-2 Land Cover Explorer](https://livingatlas.arcgis.com/landcoverexplorer/#mapCenter=-158.007%2C21.395%2C11&mode=step&timeExtent=2017%2C2022&year=2022)
- [Unleash the Power of 3D: Crafting Stunning Land Cover Maps with R by Milos Makes Maps](https://www.youtube.com/watch?v=y_Kzg24Ciuo)